---
title: "K-means Clustering with smoothclust"
author: "Lauren Rusnak"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages

```{r}
library(SpatialExperiment)
library(STexampleData)
library(scran)
library(ggspavis)
library(scater)
library(smoothclust)
library(stats)
```



### Load Data

*Download data object*
```{r}
spe <- Visium_humanDLPFC() # human dorsolateral prefrontal cortex dataset
spe <- spe[, colData(spe)$in_tissue == 1] # selecting spatial coodinates that are in the tissue
```

### Normalization and Feature Selection

*Calculate highly variable genes (HVGs)*
```{r}
# Normalization - transform assay counts to normalize data (Using library size factors)
spe <- logNormCounts(spe) 

is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name) # filter for mitochondrial genes (*Note grepl has 'L' at end not '1')
spe <- spe[!is_mito, ] # remove mitochondrial genes

# keep full object for plotting
spe_full <- spe
```


```{r}
# library(scran)

# fit mean-variance relationship
dec <- modelGeneVar(spe) # model the variance of the log-expression profiles for each gene
top_hvgs <- getTopHVGs(dec, prop = 0.1) # top 10% of genes with the highest variance of log-expression are chosen
spe <- spe[top_hvgs, ]
dim(spe)
```


_________________________________________________________________________


### **RUN SMOOTHCLUST**

#### Method: Uniform

```{r}
# library(smoothclust)

# run smoothclust
spe <- smoothclust(spe, method = "uniform")
```

```{r}
# check
assayNames(spe)
```


##### Dimensionality reduction

```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
```

```{r}
# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


```{r}
plotDimRed(spe, type = "PCA")
```

```{r}
pcs <- reducedDim(spe, "PCA")
plot(pcs)
```


##### Clustering

*k-means*

```{r}
Wss <- c()

for(i in 1:10){
  km <- kmeans(pcs, i)
  Wss <- c(Wss, km$tot.withinss)
}

plot(c(1:10), Wss, type = 'l') ## kink at 2 (maybe another at 4 and maybe abother at 6)

```

```{r}
km.2 <- kmeans(pcs, centers = 2)

plot(pcs[,1], pcs[,2])
for(i in 1:2){
    points(pcs[which(km.2$cluster == i),1],
          pcs[which(km.2$cluster == i),2],col=i)
  }
```




```{r}
set.seed(123)
km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

# plot(pcs[,1], pcs[,2])
# for(i in 1:7){
    # points(pcs[which(km.7$cluster == i),1],
         # pcs[which(km.7$cluster == i),2],col=i)
#  }
```

```{r}
# kmeans cluster assignments
km.clus <- km.7$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus)
```


##### Visualizations

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.8)
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors",
          size = 0.7) # annotate - for clustering labels, use smoothclust labels later
```


Reduced Dimensionality Cluster Visualization

```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



```{r}
# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```

_________________________________________________________________________


#### Method: Kernel, bandwidth = 0.02

```{r}
# library(smoothclust)

# run smoothclust
spe <- smoothclust(spe, method = "kernel", bandwidth = 0.02)
```

```{r}
# check
assayNames(spe)
```


##### Dimensionality reduction

```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
```

```{r}
# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


```{r}
plotDimRed(spe, type = "PCA")
```

```{r}
pcs <- reducedDim(spe, "PCA")
# plot(pcs)
```



##### Clustering

*k-means*

```{r}
Wss <- c()

for(i in 1:10){
  km <- kmeans(pcs, i)
  Wss <- c(Wss, km$tot.withinss)
}

plot(c(1:10), Wss, type = 'l') ## kink at 2 (maybe another at 4 and maybe abother at 6)

```

```{r}
km.2 <- kmeans(pcs, centers = 2)

# plot(pcs[,1], pcs[,2])
# for(i in 1:2){
    # points(pcs[which(km.2$cluster == i),1],
         # pcs[which(km.2$cluster == i),2],col=i)
#  }
```




```{r}
set.seed(123)
km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

# plot(pcs[,1], pcs[,2])
# for(i in 1:7){
    # points(pcs[which(km.7$cluster == i),1],
          # pcs[which(km.7$cluster == i),2],col=i)
#  }
```

```{r}
# kmeans cluster assignments
km.clus <- km.7$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus)
```


##### Visualizations

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.8)
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors",
          size = 0.7) # annotate - for clustering labels, use smoothclust labels later
```


Reduced Dimensionality Cluster Visualization

```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



```{r}
# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



_______________________________________________________________________


#### Method: kernel, bandwidth = 0.04


```{r}
# library(smoothclust)

# run smoothclust
spe <- smoothclust(spe, method = "kernel", bandwidth = 0.04)
```

```{r}
# check
assayNames(spe)
```


##### Dimensionality reduction

```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
```

```{r}
# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


```{r}
plotDimRed(spe, type = "PCA")
```

```{r}
pcs <- reducedDim(spe, "PCA")
# plot(pcs)
```



##### Clustering

*k-means*

```{r}
Wss <- c()

for(i in 1:10){
  km <- kmeans(pcs, i)
  Wss <- c(Wss, km$tot.withinss)
}

plot(c(1:10), Wss, type = 'l') ## kink at 2 (maybe another at 4 and maybe abother at 6)

```

```{r}
km.4 <- kmeans(pcs, centers = 4)

# plot(pcs[,1], pcs[,2])
# for(i in 1:2){
    # points(pcs[which(km.2$cluster == i),1],
         # pcs[which(km.2$cluster == i),2],col=i)
#  }
```




```{r}
set.seed(123)
km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

# plot(pcs[,1], pcs[,2])
# for(i in 1:7){
    # points(pcs[which(km.7$cluster == i),1],
          # pcs[which(km.7$cluster == i),2],col=i)
#  }
```

```{r}
# kmeans cluster assignments
km.clus <- km.4$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus)
```


##### Visualizations

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.8)
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors",
          size = 0.7) # annotate - for clustering labels, use smoothclust labels later
```



_______________________________________________________________________

#### Method: Kernel , bandwidth = 0.05

```{r}
# library(smoothclust)

# run smoothclust
spe <- smoothclust(spe, method = "kernel", bandwidth = 0.05)
```

```{r}
# check
assayNames(spe)
```


##### Dimensionality reduction

```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
```

```{r}
# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


```{r}
plotDimRed(spe, type = "PCA")
```

```{r}
pcs <- reducedDim(spe, "PCA")
# plot(pcs)
```



##### Clustering

*k-means*

```{r}
Wss <- c()

for(i in 1:10){
  km <- kmeans(pcs, i)
  Wss <- c(Wss, km$tot.withinss)
}

plot(c(1:10), Wss, type = 'l') ## kink at 2 (maybe another at 4 and maybe abother at 6)

```

```{r}
km.2 <- kmeans(pcs, centers = 2)

# plot(pcs[,1], pcs[,2])
# for(i in 1:2){
    # points(pcs[which(km.2$cluster == i),1],
          # pcs[which(km.2$cluster == i),2],col=i)
#  }
```




```{r}
set.seed(123)
km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

# plot(pcs[,1], pcs[,2])
# for(i in 1:7){
   # points(pcs[which(km.7$cluster == i),1],
         # pcs[which(km.7$cluster == i),2],col=i)
#  }
```

```{r}
# kmeans cluster assignments
km.clus <- km.7$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus)
```


##### Visualizations

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.8)
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors",
          size = 0.7) # annotate - for clustering labels, use smoothclust labels later
```


Reduced Dimensionality Cluster Visualization

```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```


```{r}
# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



_____________________________________________________________________



#### Method: Kernel , bandwidth = 0.1

```{r}
# library(smoothclust)

# run smoothclust
spe <- smoothclust(spe, method = "kernel", bandwidth = 0.1)
```

```{r}
# check
assayNames(spe)
```


##### Dimensionality reduction

```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
```

```{r}
# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


```{r}
plotDimRed(spe, type = "PCA")
```

```{r}
pcs <- reducedDim(spe, "PCA")
# plot(pcs)
```



##### Clustering

*k-means*

```{r}
Wss <- c()

for(i in 1:10){
  km <- kmeans(pcs, i)
  Wss <- c(Wss, km$tot.withinss)
}

plot(c(1:10), Wss, type = 'l') ## kink at 2 (maybe another at 4 and maybe abother at 6)

```

```{r}
km.2 <- kmeans(pcs, centers = 2)

plot(pcs[,1], pcs[,2])
for(i in 1:2){
    points(pcs[which(km.2$cluster == i),1],
          pcs[which(km.2$cluster == i),2],col=i)
  }
```




```{r}
set.seed(123)
km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

plot(pcs[,1], pcs[,2])
for(i in 1:7){
    points(pcs[which(km.7$cluster == i),1],
          pcs[which(km.7$cluster == i),2],col=i)
  }
```

```{r}
# kmeans cluster assignments
km.clus <- km.7$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus)
```


##### Visualizations

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.8)
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors",
          size = 0.7) # annotate - for clustering labels, use smoothclust labels later
```


Reduced Dimensionality Cluster Visualization

```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



```{r}
# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```