---
title: "ST Book Ex RR"
author: "Lauren Rusnak"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data (5)

```{r include=FALSE}
library(SpatialExperiment)
library(STexampleData)
```


```{r}
# load object
spe <- Visium_humanDLPFC()
```

```{r}
# check object
spe
```

```{r}
# number of genes (rows) and spots (columns)
dim(spe)
```

```{r}
# names of 'assays'
assayNames(spe)
```

```{r}
# row (gene) data
head(rowData(spe))
```

```{r}
# column (spot) data
head(colData(spe))
```

```{r}
# spatial coordinates
head(spatialCoords(spe))
```

```{r}
# image data
imgData(spe)
```
  + *What do the columns output by imgData(spe) mean?*



_________________________________________________________________



## Quality COntrol - 6


```{r}
library(ggspavis)
```

```{r}
# plot spatial coordinates (spots)
plotSpots(spe)
```


```{r}
library(scater)
```

```{r}
# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]
dim(spe)
```


*Mitochondrial Genes*
```{r}
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)
```

```{r}
rowData(spe)$gene_name[is_mito]
```

*Cells per spot*
```{r}
# calculate per-spot QC metrics (cell_count) and store in colData
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
head(colData(spe))
```


```{r}
# histogram of library sizes
hist(colData(spe)$sum, breaks = 20)
```

  + Library size = total sum of UMI counts per spot
  + The distribution is relatively smooth, and there are no obvious issue such as a spike at very low library sizes.


```{r}
# plot library size vs. number of cells per spot
plotQC(spe, type = "scatter", 
       metric_x = "cell_count", metric_y = "sum", 
       threshold_y = 600)
```

  + The plot shows that setting a filtering threshold for library size (e.g. at the value shown) does not appear to select for any obvious biologically consistent group of spots.
  


```{r}
# select QC threshold for library size (600 is arbitaray in this example)
qc_lib_size <- colData(spe)$sum < 600
table(qc_lib_size)
```

```{r}
colData(spe)$qc_lib_size <- qc_lib_size
```

```{r}
# check spatial pattern of discarded spots
plotQC(spe, type = "spots", 
       discard = "qc_lib_size")
```

  + Want to see NO obvious spatial pattern that correlates with known biological features
  + Above satisfies, threshold good
  + Below violates, threshold bad (too high)
  + *What is the threshold for?*

```{r}
# check spatial pattern of discarded spots if threshold is too high
qc_lib_size_2000 <- colData(spe)$sum < 2000
colData(spe)$qc_lib_size_2000 <- qc_lib_size_2000
plotQC(spe, type = "spots", 
       discard = "qc_lib_size_2000")
```

```{r}
# plot ground truth (manually annotated) layers
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors")
```


*Number if Expressed Features*
```{r}
# histogram of numbers of expressed genes
hist(colData(spe)$detected, breaks = 20)
```

```{r}
# plot number of expressed genes vs. number of cells per spot
plotQC(spe, type = "scatter", 
       metric_x = "cell_count", metric_y = "detected", 
       threshold_y = 400)
```

```{r}
# select QC threshold for number of expressed genes
qc_detected <- colData(spe)$detected < 400
table(qc_detected)
```

```{r}
colData(spe)$qc_detected <- qc_detected

# check spatial pattern of discarded spots
plotQC(spe, type = "spots", 
       discard = "qc_detected")
```

```{r}
# check spatial pattern of discarded spots if threshold is too high
qc_detected_1000 <- colData(spe)$detected < 1000
colData(spe)$qc_detected_1000 <- qc_detected_1000
plotQC(spe, type = "spots", 
       discard = "qc_detected_1000")
```



*Proportion of mitochondrial reads*
```{r}
# histogram of mitochondrial read proportions
hist(colData(spe)$subsets_mito_percent, breaks = 20)
```

```{r}
# plot mitochondrial read proportion vs. number of cells per spot
plotQC(spe, type = "scatter", 
       metric_x = "cell_count", metric_y = "subsets_mito_percent", 
       threshold_y = 28)
```

```{r}
# select QC threshold for mitochondrial read proportion
qc_mito <- colData(spe)$subsets_mito_percent > 28
table(qc_mito)
```

```{r}
colData(spe)$qc_mito <- qc_mito

# check spatial pattern of discarded spots
plotQC(spe, type = "spots", 
       discard = "qc_mito")
```

```{r}
# check spatial pattern of discarded spots if threshold is too high
qc_mito_25 <- colData(spe)$subsets_mito_percent > 25
colData(spe)$qc_mito_25 <- qc_mito_25
plotQC(spe, type = "spots", 
       discard = "qc_mito_25")
```



*Number of cells per spot*
```{r}
# histogram of cell counts
hist(colData(spe)$cell_count, breaks = 20)
```

```{r}
# distribution of cells per spot
tbl_cells_per_spot <- table(colData(spe)$cell_count)

# plot number of expressed genes vs. number of cells per spot
plotQC(spe, type = "scatter", 
       metric_x = "cell_count", metric_y = "detected", 
       threshold_x = 10)
```

```{r}
# select QC threshold for number of cells per spot
qc_cell_count <- colData(spe)$cell_count > 10
table(qc_cell_count)
```

```{r}
colData(spe)$qc_cell_count <- qc_cell_count

# check spatial pattern of discarded spots
plotQC(spe, type = "spots", 
       discard = "qc_cell_count")
```



*Remove low-quality spots*
```{r}
# number of discarded spots for each metric
apply(cbind(qc_lib_size, qc_detected, qc_mito, qc_cell_count), 2, sum)
```

```{r}
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
table(discard)
```

```{r}
# store in object
colData(spe)$discard <- discard

# check spatial pattern of combined set of discarded spots
plotQC(spe, type = "spots", 
       discard = "discard")
```

```{r}
# remove combined set of low-quality spots
spe <- spe[, !colData(spe)$discard]
dim(spe)
```








## Normalization - 7

Previous steps
```{r}
# QUALITY CONTROL (QC) ~ Previous steps

library(scater)
# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
# select QC thresholds
qc_lib_size <- colData(spe)$sum < 600
qc_detected <- colData(spe)$detected < 400
qc_mito <- colData(spe)$subsets_mito_percent > 28
qc_cell_count <- colData(spe)$cell_count > 10
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]
```


```{r}
library(scran)
```

```{r}
# calculate library size factors
spe <- computeLibraryFactors(spe)

summary(sizeFactors(spe))
```

```{r}
hist(sizeFactors(spe), breaks = 20)
```

```{r}
# calculate logcounts and store in object
spe <- logNormCounts(spe)

# check
assayNames(spe)
```

```{r}
dim(counts(spe))
```

```{r}
dim(logcounts(spe))
```




## Feature Selection - 8

*HVGs*
```{r}
# remove mitochondrial genes
spe <- spe[!is_mito, ]
dim(spe)
```

```{r}
library(scran)

# fit mean-variance relationship
dec <- modelGeneVar(spe)

# visualize mean-variance relationship
fit <- metadata(dec)
plot(fit$mean, fit$var, 
     xlab = "mean of log-expression", ylab = "variance of log-expression")
curve(fit$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)
```

```{r}
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
length(top_hvgs)
```


## Dimensionality reduction - 9

previous steps
```{r}
# LOAD DATA

library(SpatialExperiment)
library(STexampleData)
spe <- Visium_humanDLPFC()

# QUALITY CONTROL (QC)

library(scater)
# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
# select QC thresholds
qc_lib_size <- colData(spe)$sum < 600
qc_detected <- colData(spe)$detected < 400
qc_mito <- colData(spe)$subsets_mito_percent > 28
qc_cell_count <- colData(spe)$cell_count > 10
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]

# NORMALIZATION

library(scran)
# calculate logcounts using library size factors
spe <- logNormCounts(spe)

# FEATURE SELECTION

# remove mitochondrial genes
spe <- spe[!is_mito, ]
# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
```


*PCA*
```{r}
# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs)

reducedDimNames(spe)
```

```{r}
dim(reducedDim(spe, "PCA"))
```

*Uniform Manifold Approximation and Projection (UMAP)*
```{r}
# compute UMAP on top 50 PCs
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")

reducedDimNames(spe)
```

```{r}
dim(reducedDim(spe, "UMAP"))
```

```{r}
# update column names for easier plotting
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


*Visualizations*
```{r}
library(ggspavis)
```

```{r}
# plot top 2 PCA dimensions
plotDimRed(spe, type = "PCA")
```

```{r}
# plot top 2 UMAP dimensions
plotDimRed(spe, type = "UMAP")
```




## Clustering - 10

previous steps
```{r}
# LOAD DATA

library(SpatialExperiment)
library(STexampleData)
spe <- Visium_humanDLPFC()

# QUALITY CONTROL (QC)

library(scater)
# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
# select QC thresholds
qc_lib_size <- colData(spe)$sum < 600
qc_detected <- colData(spe)$detected < 400
qc_mito <- colData(spe)$subsets_mito_percent > 28
qc_cell_count <- colData(spe)$cell_count > 10
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]

# NORMALIZATION

library(scran)
# calculate logcounts using library size factors
spe <- logNormCounts(spe)

# FEATURE SELECTION

# remove mitochondrial genes
spe <- spe[!is_mito, ]
# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)

# DIMENSIONALITY REDUCTION

# compute PCA
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs)
# compute UMAP on top 50 PCs
set.seed(123)
spe <- runUMAP(spe, dimred = "PCA")
# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
```


*Non-spatial clustering on HVGs*
```{r}
# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)
```

```{r}
# store cluster labels in column 'label' in colData
colLabels(spe) <- factor(clus)
```


```{r}
library(ggspavis)
```

```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors")
```

```{r}
# plot ground truth labels in spatial coordinates
plotSpots(spe, annotate = "ground_truth", 
          palette = "libd_layer_colors")
```

```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors")
```

```{r}
# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors")
```


